Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-09-12 Manuscript.qmd
Added time warped graphs into supplementary and changed dtw function to provide standard deviation for warped series

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

@@

Recent research particularly of *Patella* spp. in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io].

In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which allows navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio. By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with research background information can be found in Table [1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not allow them to be re-analysed.

```{r SST_Data}

# Read the CSV files, skipping the comments
Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")

@@

@@

@@

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as LA-ICP-MS [@Graniero2017-io]. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl.
#| fig-height: 3
#| fig-width: 6
#| label: fig-SSTs

# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SST (°C)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))

@@

We also studied 7 archaeological specimens of *P. vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates. That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.

@@

@@

High-quality $\delta$^18^O values were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaurata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found. In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(paste(delta^{18},"O (\u2030)")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively). Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp. shells living 2–3 years and *P. vulgata* living regularly over 3 years. In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e. specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.

```{r Patella_Isotope_plot,dev='png',dpi=300,fig.pos='H'}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Patella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_Surge_Godino_2024]). The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.

@@

  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data)
}

@@

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

@@

Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value of the individual *P. vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.

@@

P_individual <- P_all+ylim(0.2,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

@@

Line scans reflect well the changes previously indicated by $\delta$^18^O values and also indicate about two years of recorded growth. That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.

@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O values (similar graphs for the other *Nacella* spp. specimens can be found in the supplementaries). Compared to the *P. vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and thus a lower chance of misaligning annual extremes.

@@

| Species       | Locality       | Specimen  | R^2^                    | Study                             |     |
|------------|------------|------------|------------|------------|------------|
| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |     |
|               |                | LAN545    | 0.86                    |                                   |     |
|               |                | LAN554    | 0.78                    |                                   |     |
|               |                | LAN559    | 0.82                    |                                   |     |
| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |     |
|               |                | ISTPC2    | 0.84                    |                                   |     |
|               | Crete          | AF1911A   | 0.91                    |                                   |     |
|               |                | AF3003A   | 0.92                    |                                   |     |
|               | Israel         | AKKPC2    | 0.96                    |                                   |     |
|               |                | AKKPC3    | 0.89                    |                                   |     |
|               |                | FRMPC1    | 0.84                    |                                   |     |
|               |                | FRMPC2    | 0.96                    |                                   |     |
|               | Libya          | MO31A     | 0.83                    |                                   |     |
|               |                | **MP64A** | **0.33**                |                                   |     |
|               |                | MP67A     | 0.96                    |                                   |     |
|               |                | MP68A     | 0.81                    |                                   |     |
|               | Malta          | MA10      | 0.82                    |                                   |     |
|               | Tunisia        | TUNPC1    | 0.81                    |                                   |     |
|               |                | TUNPC2    | 0.78                    |                                   |     |
|               | Turkey         | ANTPC1    | 0.95                    |                                   |     |
|               |                | ANTPC2    | 0.93                    |                                   |     |
|               |                | KIZPC1    | 0.94                    |                                   |     |
|               |                | KIZPC2    | 0.86                    |                                   |     |
| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |     |
|               |                | JL2       | 0.8\*\*                 |                                   |     |
| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |     |
|               |                | JM30      | 0.83\*\*                |                                   |     |
| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |     |

: Correlations of Mg/Ca ratios with $\delta$^18^O values and SST found in other studies. Bold R^2^ values are interpreted as outliers. \* indicates that only SST data and no other geochemical data was used. \*\* these samples were used to determine a shared R^2^-value of 0.79.
:::

>>>>

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline and *Patella caerulea* in the Mediterranean have repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these two species as well as *Nacella magellanica* together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

source(here("data","IsoError.R"))
IsoError <- IsoError %>% 
  arrange(sampleID)

@@

Recent research particularly of *Patella* spp. in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io]. Using modern reference material Graniero *et al*. [-@Granieo2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios. This study included two species also used in our study (*Patella vulgata*, *Nacella deaurata*), with one *P. vulgata* specimen (i.e. specimen ORK-LT5) being analysed by both. We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using \$\\delta\$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].

In this study, we repeat and expand upon the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio. By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution \$\\delta\$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with research background information can be found in Table [1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail from the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not allow them to be re-analysed.

```{r SS_Data}

# Read the CSV files, skipping the comments
Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
Westray_SSS <- read_csv(here("data","Westray-SSS.csv"), comment = "#")
TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
TdF_SSS <- read_csv(here("data","Tierra_del_Fuego-SSS.csv"), comment = "#")

@@

# Combine the datasets with new names and add year and month columns
SSS_df <- bind_rows(Westray_SSS %>% mutate(dataset = "Westray - Orkney"),
                         TdF_SSS %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
  mutate(analysed_sss = so,
         time = as_datetime(time),
         year = year(time),
         month = month(time, label = TRUE)) %>%
  group_by(dataset, year, month) %>%
  summarise(mean_sss = mean(analysed_sss, na.rm = TRUE),
            sd_sss = sd(analysed_sss, na.rm = TRUE))

@@

WestrayMin <- SSS_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sss) %>% min() %>% round(1)
WestrayMax <- SSS_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sss) %>% max() %>% round(1)

CamBayMin <- SSS_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sss) %>% min() %>% round(1)
CamBayMax <- SSS_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sss) %>% max() %>% round(1)

@@

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for \$\\delta\$^18^O and \$\\delta\$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io]. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023, monthly sea surface salinity values from 1993 to 2021. Values were sourced from @Good2020-nl.
#| fig-height: 3
#| fig-width: 6
#| label: fig-SSTs

# Plot the data
SST_plot <- SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SST (°C)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))

# Plot the data
SSS_plot <- SSS_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sss = sd(mean_sss, na.rm = TRUE),
            mean_sss = mean(mean_sss, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sss, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sss - sd_sss, ymax = mean_sss + sd_sss,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SSS (so)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = "none")+  
  scale_y_continuous(breaks = c(32,33, 34,35,36),limits = c(32,36))

SST_plot +SSS_plot

@@

We also studied 7 archaeological specimens of *P. vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba].

@@

High-quality \$\\delta\$^18^O values were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaurata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found. In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells. The exact location of the sampled carbonate powder is not available apart from the distance to the growth edge, where reported. Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_errorbar(aes(ymin = d18O - 0.1, ymax = d18O + 0.1), width = 0.2) +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(paste(delta^{18},"O (\u2030)")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential \$\\delta\$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively). Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp. shells living 2–3 years and *P. vulgata* living regularly over 3 years. In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e. specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7. Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates. That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.

```{r Patella_Isotope_plot,dev='png',dpi=300,fig.pos='H'}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species)) #%>%

Patella_iso <- Patella_iso %>% 
  arrange(sampleID) %>%
  bind_cols(d18Osd=IsoError$d18Osd)

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Patella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_errorbar(aes(ymin = d18O - d18Osd, ymax = d18O + d18Osd), width = 0.2) +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

###

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of \$\\delta\$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This can involve selecting appropriate distance measures and constraints to ensure meaningful alignment. We used the default settings for symmetric step patterns and euclidean distance measures (see code in [@Hausmann_Surge_Godino_2024]).

@@

  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data) %>% 
    arrange(x)
}

write_csv(Corr_data,here("data","Corr_data.csv"))

@@

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its \$\\delta\$^18^O values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

@@

Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (\$\\delta\$^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value of the individual *P. vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.

@@

P_individual <- P_all+ylim(0.2,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  stat_cor(
    aes(label = paste("P-value: ", ..p..)),
    label.x = 0.2, label.y = 0.6, 
    size = 6
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

@@

Line scans reflect well the changes previously indicated by \$\\delta\$^18^O values and also indicate about two years of recorded growth. That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.

@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their \$\\delta\$^18^O values (similar graphs for the other *Nacella* spp. specimens can be found in the supplementaries). Compared to the *P. vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and thus a lower chance of misaligning annual extremes.

@@

| Species          | Locality         | Specimen   | R^2^                    | Study                             |     |
|------------|------------|------------|------------|------------|------------|
| *P. depressa*    | Northern Spain   | LAN541     | 0.87                    | [@Garcia-Escarzaga2021-ij]        |     |
|                  |                  | LAN545     | 0.86                    |                                   |     |
|                  |                  | LAN554     | 0.78                    |                                   |     |
|                  |                  | LAN559     | 0.82                    |                                   |     |
| *P. caerulea*    | Croatia          | ISTPC1     | 0.9                     | [@Hausmann2019-fi]                |     |
|                  |                  | ISTPC2     | 0.84                    |                                   |     |
|                  | Crete            | AF1911A    | 0.91                    |                                   |     |
|                  |                  | AF3003A    | 0.92                    |                                   |     |
|                  | Israel           | AKKPC2     | 0.96                    |                                   |     |
|                  |                  | AKKPC3     | 0.89                    |                                   |     |
|                  |                  | FRMPC1     | 0.84                    |                                   |     |
|                  |                  | FRMPC2     | 0.96                    |                                   |     |
|                  | Libya            | MO31A      | 0.83                    |                                   |     |
|                  |                  | **MP64A**  | **0.33**                |                                   |     |
|                  |                  | MP67A      | 0.96                    |                                   |     |
|                  |                  | MP68A      | 0.81                    |                                   |     |
|                  | Malta            | MA10       | 0.82                    |                                   |     |
|                  | Tunisia          | TUNPC1     | 0.81                    |                                   |     |
|                  |                  | TUNPC2     | 0.78                    |                                   |     |
|                  | Turkey           | ANTPC1     | 0.95                    |                                   |     |
|                  |                  | ANTPC2     | 0.93                    |                                   |     |
|                  |                  | KIZPC1     | 0.94                    |                                   |     |
|                  |                  | KIZPC2     | 0.86                    |                                   |     |
| *P. rustica*     | Gibraltar        | **JL1**    | **0.02**                | [@Ferguson2011-zl]                |     |
|                  |                  | JL2        | 0.8\*\*                 |                                   |     |
| *P. caerulea*    | Gibraltar        | JM00       | 0.69\*\*                |                                   |     |
|                  |                  | JM30       | 0.83\*\*                |                                   |     |
| *P. vulgata*     | Orkney           | ORK-LT5    | not reported, here 0.88 | [@Graniero2017-io] and this study |     |
|                  |                  | QG1-7188-1 | 0.88                    | this study                        |     |
|                  |                  | QG1-7189-2 | 0.95                    |                                   |     |
|                  |                  | QG1-7246-1 | 0.81                    |                                   |     |
|                  |                  | QG2-1061-1 | 0.90                    |                                   |     |
|                  |                  | QG2-1064-1 | 0.90                    |                                   |     |
|                  |                  | QG2-7180-1 | 0.87                    |                                   |     |
|                  |                  | QG2-7180-2 | 0.90                    |                                   |     |
| *N. deaurata*    | Tierra del Fuego | ND-1016-3  | 0.95                    |                                   |     |
|                  |                  | ND-1016-4  | 0.76                    |                                   |     |
| *N. magellanica* | Tierra del Fuego | NM-1016-1  | 0.92                    |                                   |     |
|                  |                  | NM-1016-3  | 0.84                    |                                   |     |

: Correlations of Mg/Ca ratios with \$\\delta\$^18^O values and SST found in other studies. Bold R^2^ values are interpreted as outliers. \* indicates that only SST data and no other geochemical data was used. \*\* these samples were used to determine a shared R^2^-value of 0.79.
:::
